---
title: "Analysis results"
author: "Robby Butarbutar"
date: "2024-08-27"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Load packages**

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(openxlsx)
library(dplyr)
library(plyr)
library(vegan)
library(ggplot2)
library(gridExtra)
library(pairwiseAdonis)
library(indicspecies)
library(geometry)
library(plotly)
```

*Analysis across habitats*

**Reading and preparing data**

The abundace data is pooled by landscapes, transects and clusters (Habitat)

```{r}
# Read in the Excel file
df <- read.xlsx("C:/Users/Robby Butarbutar/OneDrive - University of Kent/PhD/Field Data/bird_survey_sulawesi_2024.xlsx", 
                  sheet = "PC_v3")
  
# Remove swift (Apodidae)
  df_filtered <- df %>% filter(Family != "Apodidae")

# Endemic/non-endemic dataset
  df_filtered <- df_filtered %>% filter(ENDEMIC == "NO")
  
  #pooled by habitat clusters
    df_transformed <- df_filtered %>%
      group_by(LANDSCAPE, TRANSECT, CLUSTER, HABITAT, ENGLISH_NAME) %>%
      dplyr::summarise(Total_count = sum(COUNT, na.rm = TRUE)) %>%
      pivot_wider(names_from = ENGLISH_NAME, values_from = Total_count, values_fill = 0)
    # Set rownames as POINTID
    df_transformed$row.id <- paste(df_transformed$LANDSCAPE, df_transformed$TRANSECT,
                                   df_transformed$CLUSTER, sep = "_")
    df_transformed <- column_to_rownames(df_transformed, var = "row.id")
    df_final <- df_transformed[,-c(1:4)]
    
# Log transformation
  sp_site <- log1p(df_final)
    
```


**Non-metric Multidimensional Scale (NMDS)**

```{r}
# Create a distance matrix
data_dist <- vegdist(sp_site, method = "bray")

# Perform NMDS
NMDS <- metaMDS(data_dist, trymax =250, k = 3)

# Extract NMDS coordinates and create a data frame
NMDS_df <- merge(data.frame(NMDS$points),
                 df_transformed[,c("LANDSCAPE", "HABITAT")], 
                 by.x = "row.names",by.y = "row.names", all.x = T)
rownames(NMDS_df) <- NMDS_df$Row.names
NMDS_df <- NMDS_df[, -1]
names(NMDS_df)[4:5] <- c("site", "habitat")

NMDS

```

**Visualise NMDS points**

```{r}
# Function to calculate convex hull for 3D points
compute_convex_hull <- function(df) {
  ch <- convhulln(as.matrix(df[, c("MDS1", "MDS2", "MDS3")]))
  df[ch, ]
}

# Calculate convex hulls for each habitat
hull_Coconut <- compute_convex_hull(NMDS_df[NMDS_df$habitat == "Coconut", ])
hull_Coconut_near_forest <- compute_convex_hull(NMDS_df[NMDS_df$habitat == "Coconut near forest", ])
hull_Forest <- compute_convex_hull(NMDS_df[NMDS_df$habitat == "Forest", ])

# 3D Plot with convex hulls
plot_3d <- plot_ly(NMDS_df, x = ~MDS1, y = ~MDS2, z = ~MDS3, 
                   color = ~habitat, colors = c("Coconut" = "#800080", 
                                                "Forest" = "olivedrab", 
                                                "Coconut near forest" = "#1E90FF"),
                   #symbol = ~as.factor(site), symbols = c(6, 19, 17, 10), 
                   marker = list(size = 5)) %>%
  add_markers() %>%
  add_trace(data = hull_Coconut, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, facecolor = rep("#800080", 36), name = "Coconut") %>%
  add_trace(data = hull_Coconut_near_forest, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, facecolor = rep("#1E90FF", 36), name = "Coconut near forest") %>%
  add_trace(data = hull_Forest, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, facecolor = rep("olivedrab", 36), name = "Forest") %>%
  layout(scene = list(xaxis = list(title = 'MDS1'),
                      yaxis = list(title = 'MDS2'),
                      zaxis = list(title = 'MDS3'),
                      legend = list(title = list(text = 'Habitat type'))))
# Show plot
plot_3d
```

**Adonis**

```{r}
#ADONIS 

adon2<-adonis2(df_final ~ HABITAT, data=df_transformed)
adon2
pairwise_results <- pairwise.adonis(df_final, df_transformed$HABITAT)
print(pairwise_results)
```

**Indicator species analysis**

```{r}
##indicatior species analysis
indval<-multipatt(sp_site, df_transformed$HABITAT, func="IndVal.g", duleg=F,
                  #restcomb = c(3,4),
                  control=how(nperm=999))

summary(indval)

```


*Analysis across crop setting in coconut farms (BY CLUSTER)*

**Reading and preparing data**

```{r}
# Read in the Excel file
df <- read.xlsx("C:/Users/Robby Butarbutar/OneDrive - University of Kent/PhD/Field Data/bird_survey_sulawesi_2024_9SEP24.xlsx", 
                  sheet = "PC_v3")
  
# Remove swift (Apodidae)
  df_filtered <- df %>% filter(Family != "Apodidae")
  
  #subset data coconut only
  #df_filtered <- df_filtered %>% filter(HABITAT != "Forest")
  
  # Endemic/non-endemic dataset
  #df_filtered <- df_filtered %>% filter(ENDEMIC == "NO")
    
  # Transform into appropriate format for analysis
  # pooled by crop setting
  df_transformed <- df_filtered %>%
    group_by(LANDSCAPE, TRANSECT, CLUSTER, Crop_setting, ENGLISH_NAME) %>%
    dplyr::summarise(Total_count = sum(COUNT, na.rm = TRUE)) %>%
    pivot_wider(names_from = ENGLISH_NAME, values_from = Total_count, values_fill = 0)
  # Set rownames as POINTID
  df_transformed$row.id <- paste(substr(df_transformed$LANDSCAPE, 1, 2), 
                                 df_transformed$TRANSECT,
                                 df_transformed$CLUSTER,
                                 substr(df_transformed$Crop_setting, 1, 2),
                                 sep = "_")
  df_transformed <- column_to_rownames(df_transformed, var = "row.id")
    # merge column crop setting and cluster
  df_transformed$CLUSTER <- ifelse(df_transformed$CLUSTER == 2, "NF", "")
  df_transformed$CROP_SETTING <- ifelse(df_transformed$CLUSTER == "NF", 
                                        paste(df_transformed$Crop_setting, df_transformed$CLUSTER), 
                                 df_transformed$CROP_SETTING <- df_transformed$Crop_setting)
  df_transformed <- df_transformed[, -which(names(df_transformed) %in% c("CLUSTER"))]
  #df_transformed <- column_to_rownames(df_transformed, var = "POINTID")
  df_final <- df_transformed[,-which(names(df_transformed) %in% c("CROP_SETTING", "LANDSCAPE", "TRANSECT", "Crop_setting"))]
  
  # Log transformation
  sp_site <- log1p(df_final)
  

```


**Non-metric Multidimensional Scale (NMDS)**

```{r}
# Create a distance matrix
data_dist <- vegdist(sp_site, method = "bray")

# Perform NMDS
NMDS <- metaMDS(data_dist, trymax =250, k = 3)

# Extract NMDS coordinates and create a data frame
NMDS_df <- merge(data.frame(NMDS$points),
                 df_transformed[,c("LANDSCAPE", "CROP_SETTING","Crop_setting")], 
                 by.x = "row.names",by.y = "row.names", all.x = T)
rownames(NMDS_df) <- NMDS_df$Row.names
NMDS_df <- NMDS_df[, -1]
#names(NMDS_df)[4:5] <- c("site", "crop_setting")

NMDS

```

**Visualise NMDS points**

```{r}
library(patchwork)

# Function to calculate convex hull for 2D points
compute_convex_hull_2D <- function(df, x, y) {
  ch <- chull(df[[x]], df[[y]])
  df[ch, ]
}

plot_mds <- function(MDSX, MDSY) {
# Calculate convex hulls for each habitat on different planes
# Plane: MDS1 vs MDS2
hull_MONO_12 <- compute_convex_hull_2D(NMDS_df[NMDS_df$Crop_setting == "MONO", ], MDSX, MDSY)
hull_INTERCROP_12 <- compute_convex_hull_2D(NMDS_df[NMDS_df$Crop_setting == "INTERCROP", ], MDSX, MDSY)
hull_FOREST_12 <- compute_convex_hull_2D(NMDS_df[NMDS_df$Crop_setting == "FOREST", ], MDSX, MDSY)

# Plot for MDS1 vs MDS2
plotmds <- ggplot(NMDS_df, aes(x = .data[[MDSX]], y = .data[[MDSY]], col = Crop_setting)) +
  geom_polygon(data = rbind(hull_MONO_12, hull_INTERCROP_12, hull_FOREST_12), aes(fill = Crop_setting, group = Crop_setting), alpha = 0.2) +
  geom_point(cex = 3, stroke = 0.5) +
  scale_color_manual(values = c("FOREST" = "olivedrab", "INTERCROP" = "#7050FF", "MONO" ="#654321"), name = "Habitat type") +
  scale_fill_manual(values = c("FOREST" = "olivedrab", "INTERCROP" = "#7050FF", "MONO" ="#654321"), name = "Habitat type") +
  theme_classic() + 
  theme(axis.text.x = element_text(size = 7),   
        axis.text.y = element_text(size = 7),   
        axis.title.x = element_text(size = 10),  
        axis.title.y = element_text(size = 10),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10)) 
  #labs(title = "MDS1 vs MDS2")

print(plotmds)
}

combined_plot <- (plot_mds("MDS1", "MDS2") | plot_mds("MDS1", "MDS3") | plot_mds("MDS2", "MDS3")) + plot_layout(guides = "collect", axis_titles = "keep")
combined_plot

```


# Function to calculate convex hull for 3D points
compute_convex_hull <- function(df) {
  ch <- convhulln(as.matrix(df[, c("MDS1", "MDS2", "MDS3")]))
  df[ch, ]
}


# Calculate convex hulls for each habitat (including NF)
hull_MONO_nearforest <- compute_convex_hull(NMDS_df[NMDS_df$CROP_SETTING == "MONO NF", ])
hull_MONO <- compute_convex_hull(NMDS_df[NMDS_df$CROP_SETTING == "MONO", ])
hull_INTERCROP_nearforest <- compute_convex_hull(NMDS_df[NMDS_df$CROP_SETTING == "INTERCROP NF", ])
hull_INTERCROP <- compute_convex_hull(NMDS_df[NMDS_df$CROP_SETTING == "INTERCROP", ])
hull_FOREST <- compute_convex_hull(NMDS_df[NMDS_df$CROP_SETTING == "FOREST",])

# Calculate convex hulls for each habitat
hull_MONO <- compute_convex_hull(NMDS_df[NMDS_df$Crop_setting == "MONO", ])
hull_INTERCROP <- compute_convex_hull(NMDS_df[NMDS_df$Crop_setting == "INTERCROP", ])
hull_FOREST <- compute_convex_hull(NMDS_df[NMDS_df$Crop_setting == "FOREST",])

# 3D Plot with convex hulls (including NF)
plot_3d <- plot_ly(NMDS_df, x = ~MDS1, y = ~MDS2, z = ~MDS3, 
                   color = ~CROP_SETTING, colors = c("MONO" = "#654321", "MONO NF"="#898022",
                                                "INTERCROP" = "#7050FF", "INTERCROP NF" = "#9100FF",
                                                "FOREST" = "olivedrab"),
                   #symbol = ~as.factor(CLUSTER), symbols = c(6, 19, 17), 
                   marker = list(size = 5)) %>%
  add_markers() %>%
  add_trace(data = hull_MONO, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, facecolor = rep("#654321", 54), showlegend=F, name = "Monocrop") %>%
  add_trace(data = hull_MONO_nearforest, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, facecolor = rep("#898022", 54),showlegend = F, name = "Monocrop near forest") %>%
  add_trace(data = hull_INTERCROP, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, name = "Intercrop", showlegend = F, facecolor = rep("#7050FF", 54))%>%
  add_trace(data = hull_INTERCROP_nearforest, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, name = "Intercrop near forest", showlegend = F, facecolor = rep("#9100FF", 54)) %>%
  add_trace(data = hull_FOREST, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, name = "Forest", showlegend = F, facecolor = rep("olivedrab", 54)) %>%
  layout(scene = list(xaxis = list(title = 'MDS1'),
                      yaxis = list(title = 'MDS2'),
                      zaxis = list(title = 'MDS3'),
                      legend = list(title = list(text = 'Crop setting'))))
# Show plot
plot_3d

# 3D Plot with convex hulls
plot_3d <- plot_ly(NMDS_df, x = ~MDS1, y = ~MDS2, z = ~MDS3, 
                   color = ~Crop_setting, colors = c("MONO" = "#654321",
                                                "INTERCROP" = "#7050FF",
                                                "FOREST" = "olivedrab"),
                   #symbol = ~as.factor(CLUSTER), symbols = c(6, 19, 17), 
                   marker = list(size = 5)) %>%
  add_markers() %>%
  add_trace(data = hull_MONO, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, facecolor = rep("#654321", 54), showlegend=F, name = "Monocrop") %>%
  add_trace(data = hull_INTERCROP, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, name = "Intercrop", showlegend = F, facecolor = rep("#7050FF", 54))%>%
  add_trace(data = hull_FOREST, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, name = "Forest", showlegend = F, facecolor = rep("olivedrab", 54)) %>%
  layout(scene = list(xaxis = list(title = 'MDS1'),
                      yaxis = list(title = 'MDS2'),
                      zaxis = list(title = 'MDS3'),
                      legend = list(title = list(text = 'Crop setting'))))
# Show plot
plot_3d

**Adonis**

```{r}
#ADONIS 
adon2<-adonis2(df_final ~ Crop_setting, data=df_transformed)
adon2
pairwise_results <- pairwise.adonis(df_final, df_transformed$Crop_setting)
print(pairwise_results)
```

**Indicator species analysis**

```{r}
##indicatior species analysis
indval<-multipatt(df_final, df_transformed$Crop_setting, func="r.g", duleg=F,
                  #restcomb = c(3,4),
                  control=how(nperm=999))

summary(indval)

```



*Analysis across crop setting in coconut farms (BY POINT)*

**Reading and preparing data**

```{r}
# Read in the Excel file
df <- read.xlsx("C:/Users/Robby Butarbutar/OneDrive - University of Kent/PhD/Field Data/bird_survey_sulawesi_2024_9SEP24.xlsx", 
                  sheet = "PC_v3")
  
# Remove swift (Apodidae)
  df_filtered <- df %>% filter(Family != "Apodidae")
  
  #subset data coconut only
  #df_filtered <- df_filtered %>% filter(HABITAT != "Forest")
  
  # Endemic/non-endemic dataset
  #df_filtered <- df_filtered %>% filter(ENDEMIC == "YES")
    
  # Transform into appropriate format for analysis
  # pooled by crop setting
  df_transformed <- df_filtered %>%
    group_by(LANDSCAPE, TRANSECT, CLUSTER ,POINTID, Crop_setting, ENGLISH_NAME) %>%
    dplyr::summarise(Total_count = sum(COUNT, na.rm = TRUE)) %>%
    pivot_wider(names_from = ENGLISH_NAME, values_from = Total_count, values_fill = 0)
  # Set rownames as POINTID
  df_transformed <- column_to_rownames(df_transformed, var = "POINTID")
    # merge column crop setting and cluster
  df_transformed$CLUSTER <- ifelse(df_transformed$CLUSTER == 2, "NF", "")
  df_transformed$CROP_SETTING <- ifelse(df_transformed$CLUSTER == "NF", 
                                        paste(df_transformed$Crop_setting, df_transformed$CLUSTER), 
                                 df_transformed$CROP_SETTING <- df_transformed$Crop_setting)
  df_transformed <- df_transformed[, -which(names(df_transformed) %in% c("CLUSTER"))]
  #df_transformed <- column_to_rownames(df_transformed, var = "POINTID")
  df_final <- df_transformed[,-which(names(df_transformed) %in% c("CROP_SETTING", "LANDSCAPE", "TRANSECT", "Crop_setting"))]
  
  # Log transformation
  sp_site <- log1p(df_final)
  

```


**Non-metric Multidimensional Scale (NMDS)**

```{r}
# Create a distance matrix
data_dist <- vegdist(sp_site, method = "bray")

# Perform NMDS
NMDS <- metaMDS(data_dist, trymax =250, k = 3)

# Extract NMDS coordinates and create a data frame
NMDS_df <- merge(data.frame(NMDS$points),
                 df_transformed[,c("LANDSCAPE", "CROP_SETTING","Crop_setting")], 
                 by.x = "row.names",by.y = "row.names", all.x = T)
rownames(NMDS_df) <- NMDS_df$Row.names
NMDS_df <- NMDS_df[, -1]
#names(NMDS_df)[4:5] <- c("site", "crop_setting")

NMDS

```

**Visualise NMDS points**

```{r}
# Function to calculate convex hull for 3D points
compute_convex_hull <- function(df) {
  ch <- convhulln(as.matrix(df[, c("MDS1", "MDS2", "MDS3")]))
  df[ch, ]
}


# Calculate convex hulls for each habitat (including NF)
hull_MONO_nearforest <- compute_convex_hull(NMDS_df[NMDS_df$CROP_SETTING == "MONO NF", ])
hull_MONO <- compute_convex_hull(NMDS_df[NMDS_df$CROP_SETTING == "MONO", ])
hull_INTERCROP_nearforest <- compute_convex_hull(NMDS_df[NMDS_df$CROP_SETTING == "INTERCROP NF", ])
hull_INTERCROP <- compute_convex_hull(NMDS_df[NMDS_df$CROP_SETTING == "INTERCROP", ])
hull_FOREST <- compute_convex_hull(NMDS_df[NMDS_df$CROP_SETTING == "FOREST",])

# Calculate convex hulls for each habitat
hull_MONO <- compute_convex_hull(NMDS_df[NMDS_df$Crop_setting == "MONO", ])
hull_INTERCROP <- compute_convex_hull(NMDS_df[NMDS_df$Crop_setting == "INTERCROP", ])
hull_FOREST <- compute_convex_hull(NMDS_df[NMDS_df$Crop_setting == "FOREST",])

# 3D Plot with convex hulls (including NF)
plot_3d <- plot_ly(NMDS_df, x = ~MDS1, y = ~MDS2, z = ~MDS3, 
                   color = ~CROP_SETTING, colors = c("MONO" = "#654321", "MONO NF"="#898022",
                                                "INTERCROP" = "#7050FF", "INTERCROP NF" = "#9100FF",
                                                "FOREST" = "olivedrab"),
                   #symbol = ~as.factor(CLUSTER), symbols = c(6, 19, 17), 
                   marker = list(size = 5)) %>%
  add_markers() %>%
  add_trace(data = hull_MONO, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, facecolor = rep("#654321", 54), showlegend=F, name = "Monocrop") %>%
  add_trace(data = hull_MONO_nearforest, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, facecolor = rep("#898022", 54),showlegend = F, name = "Monocrop near forest") %>%
  add_trace(data = hull_INTERCROP, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, name = "Intercrop", showlegend = F, facecolor = rep("#7050FF", 54))%>%
  add_trace(data = hull_INTERCROP_nearforest, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, name = "Intercrop near forest", showlegend = F, facecolor = rep("#9100FF", 54)) %>%
  add_trace(data = hull_FOREST, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, name = "Forest", showlegend = F, facecolor = rep("olivedrab", 54)) %>%
  layout(scene = list(xaxis = list(title = 'MDS1'),
                      yaxis = list(title = 'MDS2'),
                      zaxis = list(title = 'MDS3'),
                      legend = list(title = list(text = 'Crop setting'))))
# Show plot
plot_3d

# 3D Plot with convex hulls
plot_3d <- plot_ly(NMDS_df, x = ~MDS1, y = ~MDS2, z = ~MDS3, 
                   color = ~Crop_setting, colors = c("MONO" = "#654321",
                                                "INTERCROP" = "#7050FF",
                                                "FOREST" = "olivedrab"),
                   #symbol = ~as.factor(CLUSTER), symbols = c(6, 19, 17), 
                   marker = list(size = 5), showlegend=T) %>%
  add_markers() %>%
  add_trace(data = hull_MONO, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, facecolor = rep("#654321", 54), showlegend=F, name = "Monocrop") %>%
  add_trace(data = hull_INTERCROP, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, name = "Intercrop", showlegend = F, facecolor = rep("#7050FF", 54))%>%
  add_trace(data = hull_FOREST, x = ~MDS1, y = ~MDS2, z = ~MDS3,
            type = "mesh3d", opacity = 0.2, alphahull = 0, name = "Forest", showlegend = F, facecolor = rep("olivedrab", 54)) %>%
  layout(scene = list(xaxis = list(title = 'MDS1'),
                      yaxis = list(title = 'MDS2'),
                      zaxis = list(title = 'MDS3'),
                      legend = list(title = list(text = 'Crop setting'))))
# Show plot
plot_3d

```

**Adonis**

```{r}
#ADONIS 
adon2<-adonis2(df_final ~ Crop_setting, data=df_transformed)
adon2
pairwise_results <- pairwise.adonis(df_final, df_transformed$Crop_setting)
print(pairwise_results)
```

**Indicator species analysis**

```{r}
##indicatior species analysis
indval<-multipatt(df_final, df_transformed$Crop_setting, func="r.g", duleg=F,
                  #restcomb = c(3,4),
                  control=how(nperm=999))

summary(indval)

```

